---
title: "PNW CAR - LME for drivers of carbon accumulation rates"
author: "Katrina Poppe"
date: "`r format(Sys.time(), '%d-%B-%Y %H:%M')`"
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
    code_folding: hide
  pdf_document:
    toc: yes
    toc_depth: '3'
  word_document:
    toc: yes
    toc_depth: '3'
---

# Packages

```{r message=FALSE, warning=FALSE}

library(here) #working directory
library(tidyverse) #data manipulation
library(glmnet) #LASSO
library(stabs) #stability selection
library(lme4) #linear mixed effects model
library(performance) #model performance
library(MuMIn) #dredge
library(nlme) #random effect importance
library(glmmTMB) #marginal vs conditional R2
library(ggpubr) #R2 in plots

```

# Load data

```{r }
data_S1 <- read.csv(here("pnwcar_data.csv"))

data <- data_S1 %>%
  filter (project != "SeaGrant")

data <- data %>%
  mutate(lat_cent = scale(lat),
         z_star_cent = scale(z_star),
         RSLR_cent = scale(RSLR),
         ws_area_cent = scale(ws_area),
         rel_sed_cent = scale(rel_sed),
         dist_chan_cent = scale(dist_chan),
         wl_annual_cent = scale(wl_annual),
         wl_summer_cent = scale(wl_summer),
         wl_grow_cent = scale(wl_grow),
         sal_annual_cent = scale(sal_annual),
         sal_summer_cent = scale(sal_summer),
         sal_grow_cent = scale(sal_grow),
         ts_annual_cent = scale(ts_annual),
         ts_summer_cent = scale(ts_summer),
         ts_grow_cent = scale(ts_grow),
         tw_deep_annual_cent = scale(tw_deep_annual),
         tw_deep_summer_cent = scale(tw_deep_summer),
         tw_deep_grow_cent = scale(tw_deep_grow)
         )

data_tidal <- data %>%
  filter(land_use !="Disturbed")

```

# All sites

## LASSO variable selection

### Create predictor matrix

```{r}

variables <- c("lat_cent",
               "wl_annual_cent",
               "wl_summer_cent",
               "wl_grow_cent",
               "sal_annual_cent",
               "sal_summer_cent",
               "sal_grow_cent",
               "ts_annual_cent",
               "ts_summer_cent",
               "ts_grow_cent",
               "tw_deep_annual_cent",
               "tw_deep_summer_cent",
               "tw_deep_grow_cent")

data_model1 <- data %>%
  select(car_all, estuary, all_of(variables)) %>%
  na.omit()

# Create predictor matrix (excluding 'estuary' since LASSO handles only fixed effects)
X <- model.matrix(~ ., data_model1[, variables])[, -1]  # Remove intercept column
y <- data_model1$car_all  # Response variable

```

### Stability selection - variable importance

```{r}
set.seed(123)
stabsel_fit <- stabsel(
  x = X,
  y = y,
  fitfun = glmnet.lasso,  # LASSO fitting function built into stabs
  cutoff = 0.6,          # Variable must be selected in ≥x% of subsamples; default is 0.75
  PFER = 3,              # Controls expected false positives; default is 1
  B = 200               # Controls resampling -> more precision; default is 100
)

# Output results
print(stabsel_fit)
plot(stabsel_fit)

# Access selected variables
stable_vars <- stabsel_fit$selected
print(stable_vars)

```

### Plot variable importance
```{r}

# Extract selection proportions
selection_df <- data.frame(
  variable = names(stabsel_fit$max),
  selection_prob = stabsel_fit$max
)

# Reorder variable factor levels so highest is on top in the plot
selection_df <- selection_df %>%
  arrange(desc(selection_prob)) %>%
  mutate(variable = factor(variable, levels = rev(variable)))  # key change here

# Plot with highest proportion on top
ggplot(selection_df, aes(x = variable, y = selection_prob)) +
  geom_col(fill = "steelblue") +
  geom_hline(yintercept = stabsel_fit$cutoff, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(title = "Stability Selection: Variable Importance",
       y = "Selection Proportion",
       x = "Variable") +
  theme_minimal()

```

### Lasso model

```{r}
set.seed(123)
lasso_model <- cv.glmnet(X, y, alpha = 1)  # Alpha = 1 for LASSO (L1 penalty)
plot(lasso_model)  # Visualize cross-validation results

```

### Coefficient matrix - lambda.1se

```{r}
coef_matrix <- coef(lasso_model, s = "lambda.1se")
print(coef_matrix)

```

### Coefficient matrix - lambda.min
Used the variables selected here

```{r}
selected_vars <- names(coef(lasso_model, s = "lambda.min"))[-1]
print(selected_vars)
coef(lasso_model, s = "lambda.min")

```

## Determine best LME model

### 1st "global" model
include variables selected with LASSO

```{r warning=FALSE, message=FALSE}

data_model2 <- data %>%
  select(car_all, wl_grow_cent, sal_annual_cent, lat_cent, estuary, land_use) %>%
  na.omit()

model2 <- lmer(car_all ~
                 wl_grow_cent +
                 sal_annual_cent +
                 lat_cent +
                 (1|estuary), 
               data = data_model2,
               na.action = na.fail,
               REML = TRUE)

summary(model2)
model_performance(model2)

```

### Dredge AICc model selection

```{r}
model2_set <- dredge(model2, rank="AICc")
print(model2_set)
sw(model2_set)

```

### Importance of random effect

```{r}
model2_without <- lm(car_all ~ wl_grow_cent + sal_annual_cent + lat_cent,
                data = data_model2)
model2_with <- lmer(car_all ~ wl_grow_cent + sal_annual_cent + lat_cent + (1|estuary),
                   data = data_model2)

lr_test <- lmtest::lrtest(model2_without, model2_with)
print(lr_test)

summary(model2_without)
summary(model2_with)

```

### Compare conditional and marginal R2
If conditional (R2c) much higher than marginal (R2m), random effect contributes significantly

```{r}
model_with_RE <- glmmTMB(car_all ~ wl_grow_cent + sal_annual_cent + lat_cent + (1 | estuary), 
                         data = data_model2)

# Get marginal and conditional R2
r2_values <- r.squaredGLMM(model_with_RE)
print(r2_values)

```

### Updated model without random effect

```{r}
data_model3 <- data %>%
  select(car_all, wl_grow_cent, sal_annual_cent, lat_cent, estuary, land_use) %>%
  na.omit()

model3 <- lm(car_all ~ wl_grow_cent + sal_annual_cent + lat_cent,
              data = data_model3,
             na.action = na.fail)
summary(model3)

```

### Dredge AICc model selection - without random effect

```{r}
# Run dredge and store the model set
model3_set <- dredge(model3, rank = "AICc")
print(model3_set)

# Get variable importance
sw(model3_set)

# Extract and print adjusted R² for each model
adj_r2s <- sapply(1:nrow(model3_set), function(i) {
  mod_i <- get.models(model3_set, subset = i)[[1]]
  summary(mod_i)$adj.r.squared
})

# Print adjusted R² values
print(round(adj_r2s, 3))

```

### Model averaging
For top two models with delta AICc < 2

```{r}
# Select top 2 models (delta AICc ≤ 2)
top_models <- get.models(model3_set, subset = delta <= 2.0) 

# Perform model averaging
model_avg <- model.avg(top_models, full = TRUE)

# View the averaged coefficients and importance
summary(model_avg)
rvi <- sw(model_avg)
coefs <- summary(model_avg)$coefmat.full  # Use coefmat.full for full averaging
print(coefs)


# Create the results data frame
results_df <- as.data.frame(coefs) %>%
  rownames_to_column("Variable") %>%
  filter(Variable != "(Intercept)") %>%
  mutate(
    Lower_CI = Estimate - 1.96 * `Adjusted SE`,
    Upper_CI = Estimate + 1.96 * `Adjusted SE`,
    RVI = rvi[Variable]
  ) %>%
  select(Variable, Estimate, SE = `Adjusted SE`, Lower_CI, Upper_CI, RVI)

print(results_df)

```

### Mean adjusted R2 value for model_avg
```{r}
# Extract adjusted R² for each
adj_r2_values <- sapply(top_models, function(mod) summary(mod)$adj.r.squared)

# Average adjusted R²
mean_adj_r2 <- mean(adj_r2_values, na.rm = TRUE)
cat("Mean Adjusted R² (models used in averaging):", round(mean_adj_r2, 2), "\n")

```

### Coefficient plot model avg

```{r}

# plot coefficient values and CI
variable_mapping <- c("wl_grow_cent" = "Water table level",
                      "lat_cent" = "Latitude")

# Apply variable_mapping to the 'Variable' column in results_df
results_df$Variable <- dplyr::recode(results_df$Variable, !!!variable_mapping)

# Plot coefficient values and CI
p1 <- ggplot(results_df, aes(x = Estimate, y = Variable)) +
  geom_point(size = 1.5) +
  geom_linerange(aes(xmin = Lower_CI, xmax = Upper_CI)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    x = "Coefficient value",
    y = ""
  ) +
  theme_bw() +
  ggtitle("(a) All sites") +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5)) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 12)) +
    coord_cartesian(xlim = c(-30, 100)) +
  scale_y_discrete(labels = function(x) ifelse(x == "Placeholder", "", str_wrap(x, width = 15))) # Hide text

# Add text labels for RVI values
p1 <- p1 + geom_text(data = results_df %>% filter(!is.na(RVI)),  
                   aes(label = paste0("RVI = ", sprintf("%.1f", RVI)), 
                       x = Upper_CI + 0.02), 
                   size = 4, hjust = -0.2)
p1

```

### Plot observed vs predicted CAR - model avg without random effect

```{r warning=FALSE, message=FALSE}
data_model3$observed_values <- data_model3$car_all
data_model3$predicted_values <- predict(model_avg, type = "response")

p3 <- ggplot(data_model3, aes(x=predicted_values, y=observed_values)) +
  geom_point(aes(color = land_use), alpha = 0.8) +
  geom_abline(intercept = 0, slope=1, linetype = "dashed") +
  stat_cor(aes(label= ..rr.label..), digits = 2,  label.x = 110, label.y = 300) +
  labs(color = "Land use",
       x=expression("Predicted CAR " ~(g ~C ~m^-2 ~yr^-1)),  
       y=expression("Observed CAR " ~(g ~C ~m^-2 ~yr^-1))) +
  scale_color_manual(values = c("Reference" = "#D55E00", "Restored" = "#619CFF", "Disturbed" = "#009E73")) +
  scale_y_continuous(limits = c(-50, 360)) +
  scale_x_continuous(limits = c(-20, 150)) +
  theme_bw() +
   ggtitle("(a) All sites") +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5)) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 10),    # Adjust axis text size
        axis.title = element_text(size = 10),   # Adjust axis title size
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 10),
        legend.position = c(0.2, 0.8)) 
p3

```


# Tidal sites

## LASSO variable selection

### Create predictor matrix

```{r}
variables <- c("RSLR_cent", #tidal only
               "rel_sed_cent", #tidal only
               "ws_area_cent", #tidal only
               "dist_chan_cent", #tidal only
               "z_star_cent", #tidal only
               "lat_cent",
               "wl_annual_cent",
               "wl_summer_cent",
               "wl_grow_cent",
               "sal_annual_cent",
               "sal_summer_cent",
               "sal_grow_cent",
               "ts_annual_cent",
               "ts_summer_cent",
               "ts_grow_cent",
               "tw_deep_annual_cent",
               "tw_deep_summer_cent",
               "tw_deep_grow_cent"
               )

data_model4 <- data_tidal %>%
  select(car_all, estuary, all_of(variables)) %>%
  na.omit()

# Create predictor matrix (excluding 'estuary' since LASSO handles only fixed effects)
X <- model.matrix(~ ., data_model4[, variables])[, -1]  # Remove intercept column
y <- data_model4$car_all  # Response variable

```

### Stability selection - variable importance

```{r}
set.seed(123)
stabsel_fit <- stabsel(
  x = X,
  y = y,
  fitfun = glmnet.lasso,  # LASSO fitting function built into stabs
  cutoff = 0.6,          # Variable must be selected in ≥x% of subsamples; default is 0.75
  PFER = 3,              # Controls expected false positives; default is 1
  B = 200               # Controls resampling -> more precision; default is 100
)

# Output results
print(stabsel_fit)
plot(stabsel_fit, cex.axis = 0.8)

# Access selected variables
stable_vars <- stabsel_fit$selected
print(stable_vars)

```

### Plot variable importance

```{r}
# Extract selection proportions
selection_df <- data.frame(
  variable = names(stabsel_fit$max),
  selection_prob = stabsel_fit$max
)

# Reorder variable factor levels so highest is on top in the plot
selection_df <- selection_df %>%
  arrange(desc(selection_prob)) %>%
  mutate(variable = factor(variable, levels = rev(variable)))  # key change here

# Plot with highest proportion on top
ggplot(selection_df, aes(x = variable, y = selection_prob)) +
  geom_col(fill = "steelblue") +
  geom_hline(yintercept = stabsel_fit$cutoff, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(title = "Stability Selection: Variable Importance",
       y = "Selection Proportion",
       x = "Variable") +
  theme_minimal()

```

### Lasso model

```{r}
lasso_model <- cv.glmnet(X, y, alpha = 1)  # Alpha = 1 for LASSO (L1 penalty)
plot(lasso_model)  # Visualize cross-validation results

```

### Coefficient matrix - lambda.1se

```{r}
coef_matrix <- coef(lasso_model, s = "lambda.1se")
print(coef_matrix)

```

### Coefficient matrix - lambda.min

```{r}
selected_vars <- names(coef(lasso_model, s = "lambda.min"))[-1]
print(selected_vars)
coef(lasso_model, s = "lambda.min")

```

## Determine best LME model

### 1st "global" model
Using variables selected by LASSO
```{r warning=FALSE, message=FALSE}

data_model5 <- data_tidal %>%
  select(car_all, wl_grow_cent, estuary, land_use) %>%
  na.omit()

model5 <- lmer(car_all ~
                 wl_grow_cent +
                 (1|estuary),
               data = data_model5,
               na.action = na.fail)

summary(model5)
performance(model5)

```

### Dredge AICc model selection

```{r}
model5_set <- dredge(model5, rank="AICc")
print(model5_set)
sw(model5_set)

```

### Importance of random effect

```{r}

model5_without <- lm(car_all ~ wl_grow_cent,
                data = data_model5)
model5_with <- lmer(car_all ~ wl_grow_cent + (1|estuary),
                   data = data_model5)

lr_test <- lmtest::lrtest(model5_without, model5_with)
print(lr_test)

```

### Compare conditional and marginal R2

If conditional (R2c) much higher than marginal (R2m), random effect contributes significantly
```{r}
model_with_RE <- glmmTMB(car_all ~ wl_grow_cent + (1 | estuary), 
                         data = data_model5)

# Get marginal and conditional R2
r2_values <- r.squaredGLMM(model_with_RE)
print(r2_values)
```

### Coefficient plot

Optimal model same as 1st model
```{r}
# Extract coefficients and confidence intervals from model5
coefs_model5 <- summary(model5)$coefficients

# Create results dataframe from model5
results_df2 <- data.frame(
  Variable = rownames(coefs_model5),
  Estimate = coefs_model5[, "Estimate"],
  SE = coefs_model5[, "Std. Error"]
) %>%
  filter(Variable != "(Intercept)") %>%
  mutate(
    Lower_CI = Estimate - 1.96 * SE,
    Upper_CI = Estimate + 1.96 * SE
  )

# Add missing variables with NA to preserve space in the plot
all_vars <- c("wl_grow_cent", "lat_cent")

# Add mapping and ordering
results_df2 <- results_df2 %>%
  right_join(data.frame(Variable = all_vars), by = "Variable") %>%
  mutate(
    Variable = factor(Variable, levels = all_vars),  # correct data order
    Predictor = variable_mapping[as.character(Variable)],
    Predictor = factor(Predictor, levels = variable_mapping[all_vars])  # correct visual order
  )

# Map variable names to nice labels
variable_mapping <- c("wl_grow_cent" = "Water table level",
                      "lat_cent" = "Latitude")

results_df2$Predictor <- variable_mapping[results_df2$Variable]

# Optional: set RVI to NA or 0 since this isn't a model-averaged result
#results_df2$RVI_text <- NA
results_df2$RVI_text <- ifelse(!is.na(results_df2$Estimate), "RVI = 1.0", NA)

p2 <- ggplot(results_df2, aes(x = Estimate, y = Predictor)) +
  geom_point(size = 1.5, na.rm = TRUE) +
  geom_linerange(aes(xmin = Lower_CI, xmax = Upper_CI), na.rm = TRUE) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "Coefficient value", y = "") +
  theme_bw() +
  ggtitle("(b) Tidal sites") +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5)) +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 12)) +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 15)) +
  coord_cartesian(xlim = c(-30, 100)) 

# Add text labels for RVI values
p2 <- p2 + geom_text(data = results_df2 %>% filter(!is.na(RVI_text)),  
                   aes(label = RVI_text, x = Upper_CI + 0.02), 
                   size = 4, hjust = -0.2)
p2

```

### Plot observed vs predicted CAR

```{r warning=FALSE, message=FALSE}

data_model5$observed_values <- data_model5$car_all
data_model5$predicted_values <- predict(model5, type = "response")

p4 <- ggplot(data_model5, aes(x=predicted_values, y=observed_values)) +
  geom_point(aes(color = land_use), alpha = 0.8) +
  geom_abline(intercept = 0, slope=1, linetype = "dashed") +
  stat_cor(aes(label= ..rr.label..), digits = 2, label.x = 110, label.y = 300) +
  labs(color = "Land use",
       x=expression("Predicted CAR " ~(g ~C ~m^-2 ~yr^-1)),  
       y=expression("Observed CAR " ~(g ~C ~m^-2 ~yr^-1))) +
      scale_color_manual(values = c("Reference" = "#D55E00", "Restored" = "#619CFF")) +
  scale_y_continuous(limits = c(-50, 360)) +
  scale_x_continuous(limits = c(-20, 150)) +
  theme_bw() +
  ggtitle("(b) Tidal sites") +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5)) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 10),    # Adjust axis text size
        axis.title = element_text(size = 10),   # Adjust axis title size
          legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 10),
        legend.position = c(0.2, 0.84)) 
p4

```

# Figures for paper

## Fig S3 coefficient plot - combine all sites (p1) + tidal sites (p2)

```{r}

p2 <- p2 + theme(axis.text.y = element_blank(), 
                 axis.ticks.y = element_blank())

combined_coefficient_plot <- ggarrange(p1, p2, ncol = 2, widths = c(1.24, 1))
print(combined_coefficient_plot)

#ggsave("coefficients_combined.png", combined_coefficient_plot, width = 8, height = 4, units = "in", dpi = 300)

```

## Fig 5 obs vs pred - combine all sites (p3) + tidal sites (p4) 

```{r}
p4 <- p4 + theme(axis.title.y = element_blank())

obs_pred_combined <- ggarrange(p3, p4, ncol = 2, widths = c(1.05, 1))
print(obs_pred_combined)

#ggsave("obs_pred_combined.png", obs_pred_combined, width = 8, height = 4, units = "in", dpi = 300)

```


