---
title: "PNW CAR - ANOVAs, post-hoc pairwise tests, and regressions"
author: "Katrina Poppe"
date: "`r format(Sys.time(), '%d-%B-%Y %H:%M')`"
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
    code_folding: hide
  pdf_document:
    toc: yes
    toc_depth: '3'
  word_document:
    toc: yes
    toc_depth: '3'
---

# Packages

```{r message=FALSE, warning=FALSE}
library(here) #working directory
library(tidyverse) #data manipulation
library(rstatix) #games-howell test
library(ggpubr) #regression stats for plots

```

# Load data

```{r }
data_S1 <- read.csv(here("pnwcar_data.csv"))

data <- data_S1 %>%
  filter (project != "SeaGrant")

data_tidal <- data %>%
  filter(land_use !="Disturbed")

variable_names <- c(
  "car_all",
  "pb210_all",
  "cd",
  "set",
  "z_star",
  "RSLR",
  "lat",
  "sal_annual", "sal_summer", "sal_grow",
  "wl_annual", "wl_summer", "wl_grow",
  "ts_annual", "ts_summer", "ts_grow",
  "tw_deep_annual", "tw_deep_summer", "tw_deep_grow",
  "ws_area",
  "rel_sed",
  "dist_chan"
  )

predictor_names <- c(
  "pb210_all",
  "cd",
  "set",
  "z_star",
  "RSLR",
  "lat",
  "sal_annual", "sal_summer", "sal_grow",
  "wl_annual", "wl_summer", "wl_grow",
  "ts_annual", "ts_summer", "ts_grow",
  "tw_deep_annual", "tw_deep_summer", "tw_deep_grow",
  "ws_area",
  "rel_sed",
  "dist_chan"
  )

```

# Means by wetland type

```{r}
# CAR in g C m-2 y-1
means_car <- tapply(data$car_all, data$wetland_type2, mean, na.rm = TRUE)
sds_car <- tapply(data$car_all, data$wetland_type2, sd, na.rm = TRUE)
stats_car <- data.frame(
  wetland_type2 = names(means_car),
  mean_CAR = means_car,
  sd_CAR = sds_car
)
cat("\n--- CAR (g C m^-2 yr^-1) ---\n")
print(stats_car)

# Pb-210 Accretion rate in cm yr-1
means_pb <- tapply(data$pb210_all, data$wetland_type2, mean, na.rm = TRUE)
sds_pb <- tapply(data$pb210_all, data$wetland_type2, sd, na.rm = TRUE)
stats_pb <- data.frame(
  wetland_type2 = names(means_pb),
  mean_AR = means_pb,
  sd_AR = sds_pb
)
cat("\n--- Pb-210 (dpm g^-1) ---\n")
print(stats_pb)

# Carbon Density (CD) in mg C cm-3
means_cd <- tapply(data$cd, data$wetland_type2, mean, na.rm = TRUE)
sds_cd <- tapply(data$cd, data$wetland_type2, sd, na.rm = TRUE)
stats_cd <- data.frame(
  wetland_type2 = names(means_cd),
  mean_CD = means_cd,
  sd_CD = sds_cd
)
cat("\n--- Carbon Density (g C cm^-3) ---\n")
print(stats_cd)

```

# Welch ANOVA

```{r}
for (var_name in variable_names) {
  formula <- formula(paste(var_name, "~ wetland_type2"))
  aov_result <- oneway.test(formula, data = data, var.equal = FALSE)

  cat("Variable:", var_name, "\n")
  cat("F-value:", aov_result$statistic, "\n")
  cat("df:", aov_result$parameter, "\n")
  cat("P-Value:", aov_result$p.value, "\n")
  cat("Significant difference:", ifelse(aov_result$p.value < 0.05, "Yes", "No"), "\n")
  cat("\n")
}
```

# Games-Howell post hoc test

```{r}
for (var_name in variable_names) {
  formula <- formula(paste(var_name, "~ wetland_type2"))
  gh_result <- games_howell_test(data = data, formula, detailed = FALSE)

  # Add significance label
  gh_result <- gh_result %>%
    mutate(significance = case_when(
      p.adj < 0.05 ~ "*",                     # significant
      p.adj >= 0.05 & p.adj < 0.10 ~ "m",     # marginal
      TRUE ~ ""                               # not significant
    ))

  cat("Variable:", var_name, "\n")
  print(gh_result)
  cat("\n")
}
```

# Linear regressions
Each variable vs CAR

## All sites

```{r}
for (predictor in predictor_names) {
  formula <- as.formula(paste("car_all ~", predictor))
  model <- lm(formula, data = data)
  cat(predictor, ":\n")
  print(summary(model))
  cat("\n\n")
}
```

## Tidal sites

```{r}
for (predictor in predictor_names) {
  formula <- as.formula(paste("car_all ~", predictor))
  model <- lm(formula, data = data_tidal)
  cat(predictor, ":\n")
  print(summary(model))
  cat("\n\n")
}
```

# Figures for paper

## Fig 2

### CAR vs Pb210

```{r}
p1 <- ggplot(data, aes(x=pb210_all, y=car_all)) +
        geom_smooth(method = "lm", se=TRUE, alpha = 0.4) +
  geom_point(aes(color = land_use), alpha = 0.8) +
  stat_cor(aes(label= ..rr.label..), digits = 2,  label.x = 0.75, label.y = 320) + # if for 2 plots
  labs(color = "Land use",
       x=expression("Accretion rate " ~(cm ~yr^-1)), 
       y=expression("C accumulation rate " ~(g ~C ~m^-2 ~yr^-1))) +
  scale_color_manual(values = c("Reference" = "#D55E00", "Restored" = "#619CFF", "Disturbed" = "#009E73")) +
  ylim(-50, 370) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 10),    # Adjust axis text size
        axis.title = element_text(size = 10),   # Adjust axis title size
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 10),
        legend.position = c(0.2, 0.82)) 
p1

```
### CAR vs CD

```{r}
p2 <- ggplot(data, aes(x=cd, y=car_all)) +
        geom_smooth(method = "lm", se=TRUE, alpha = 0.4) +
  geom_point(aes(color = land_use), alpha = 0.8) +
  stat_cor(aes(label= ..rr.label..), digits = 2,  label.x = 45, label.y = 320) + # if for 2 plots
  labs(color = "Land use",
       x=expression("C density " ~(mg ~C ~cm^-3)), 
       y=expression("C accumulation rate " ~(g ~C ~m^-2 ~yr^-1))) +
  scale_color_manual(values = c("Reference" = "#D55E00", "Restored" = "#619CFF", "Disturbed" = "#009E73")) + 
  ylim(-50, 370) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 10),    # Adjust axis text size
        axis.title = element_text(size = 10),   # Adjust axis title size
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 10),
        legend.position = c(0.2, 0.8)) 
p2

```

### combine 2 plots from pb210 (p1) and CD (p2)
```{r}
p1 <- p1 + annotate("text", x = 1.0, y = -30, label = "A", size = 7, fontface = "plain") +
  theme(legend.key.height = unit(0.5, "cm"),
        axis.title = element_text(size = 11))

p2 <- p2 + theme(axis.title = element_text(size = 11),
                 axis.title.y = element_blank(),
                 legend.position = "none") +
  annotate("text", x = 55, y = -30, label = "B", size = 7, fontface = "plain")

pb210_cd_combined <- ggarrange(p1, p2, ncol = 2, widths = c(1.05, 1))
pb210_cd_combined

#ggsave("pb210_cd_combined.png", pb210_cd_combined, width = 8, height = 4, units = "in", dpi = 300)

```

## Fig 3

### CAR x Wetland type

```{r warning=FALSE}
data$wetland_type2 <- factor(data$wetland_type2, levels = c("Reference swamp", "Reference marsh", "Restored marsh", "Wet pasture", "Dry pasture"))

letters_order1 <- c("a", "a", "ab", "b", "ab")

sample_sizes <- data %>%
  filter(!is.na(car_all)) %>%
  group_by(wetland_type2) %>%
  summarise(SampleSize = n())

p3 <- ggplot(data=subset(data, !is.na(wetland_type2)), aes(x=wetland_type2, y=car_all)) +
  geom_boxplot(aes(fill=wetland_type2), show.legend=FALSE) +
  scale_fill_viridis_d(alpha = 0.5) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10)) +
  labs(
    y=expression("C accumulation rate " ~(g ~C ~m^-2 ~yr^-1)), 
    x=expression("Wetland type")) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  geom_text(data = sample_sizes, aes(label = paste(SampleSize), y = -60), size = 3) + # n at bottom
  geom_text(data = sample_sizes, aes(label = letters_order1, y = 400), size = 3) + # sig diff letters at top
  stat_summary(fun = mean, geom = "point", shape = 5, size = 2, color = "black")
p3
     
```

### Pb210 x Wetland type

``` {r warning=FALSE}
data$wetland_type2 <- factor(data$wetland_type2, levels = c("Reference swamp", "Reference marsh", "Restored marsh", "Wet pasture", "Dry pasture"))

letters_order2 <- c("ab", "a", "a", "b", "ab")

sample_sizes <- data %>%
  filter(!is.na(pb210_all)) %>%
  group_by(wetland_type2) %>%
  summarise(SampleSize = n())

p4 <- ggplot(data=subset(data, !is.na(wetland_type2)), aes(x=wetland_type2, y=pb210_all)) +
  geom_boxplot(aes(fill=wetland_type2), show.legend=FALSE) +
  scale_fill_viridis_d(alpha = 0.5) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10)) +
  labs(
    y=expression("Accretion rate " ~(cm ~yr^-1)), 
    x=expression("Wetland type")) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  geom_text(data = sample_sizes, aes(label = paste(SampleSize), y = -0.3), size = 3) + # n at bottom
  geom_text(data = sample_sizes, aes(label = letters_order2, y = 1.2), size = 3) + # sig diff letters at top
  stat_summary(fun = mean, geom = "point", shape = 5, size = 2, color = "black")
p4

```

### CD x Wetland type

```{r warning=FALSE}
data$wetland_type2 <- factor(data$wetland_type2, levels = c("Reference swamp", "Reference marsh", "Restored marsh", "Wet pasture", "Dry pasture"))

letters_order3 <- c("a", "ab", "b", "ab", "ab")

sample_sizes <- data %>%
  filter(!is.na(cd)) %>%
  group_by(wetland_type2) %>%
  summarise(SampleSize = n())

p5 <- ggplot(data=subset(data, !is.na(wetland_type2)), aes(x=wetland_type2, y=cd)) +
  geom_boxplot(aes(fill=wetland_type2), show.legend=FALSE) +
  scale_fill_viridis_d(alpha = 0.5) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10)) +
  labs(
    y=expression("C density " ~(mg ~C ~cm^-3)), 
    x=expression("Wetland type")) +
  ylim(0, 62) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  geom_text(data = sample_sizes, aes(label = paste(SampleSize), y = 0), size = 3) + # n at bottom
  geom_text(data = sample_sizes, aes(label = letters_order3, y = 62), size = 3) + # sig diff letters at top
  stat_summary(fun = mean, geom = "point", shape = 5, size = 2, color = "black")
p5

```

### combine 3 plots vertically

``` {r}
p5_blank <- p5 + theme(axis.text.x = element_blank(), axis.title.x = element_blank()) +
  annotate("text", x = 0.5, y = Inf, label = "A", size = 7, fontface = "plain", vjust = 1.5, hjust = -0.2) 
p4_blank <- p4 + theme(axis.text.x = element_blank(), axis.title.x = element_blank()) +
  annotate("text", x = 0.5, y = Inf, label = "B", size = 7, fontface = "plain", vjust = 1.5, hjust = -0.2)
p3 <- p3 +
  annotate("text", x = 0.5, y = Inf, label = "C", size = 7, fontface = "plain", vjust = 1.5, hjust = -0.2)

combined_plot <- ggpubr::ggarrange(p5_blank, p4_blank, p3, nrow = 3, heights = c(1,1,1.18), align = "v")
combined_plot

#ggsave("anova_plot_vert.png", combined_plot, width = 4.5, height = 9, dpi = 300)

```

## Fig 4

### Water level vs CAR
```{r}
fit <- lm(car_all ~ wl_grow, data = data)
r2 <- summary(fit)$r.squared
p_val <- summary(fit)$coefficients[2, 4]

r2_text <- paste0("italic(R)^2 == ", format(round(r2, 2), nsmall = 2))
p_text <- paste0("italic(p) == ", format(round(p_val, 3), nsmall = 3))

p6 <- ggplot(data, aes(x=wl_grow, y=car_all)) +
        geom_smooth(method = "lm", se=TRUE, alpha = 0.4) +
  geom_point(aes(color = land_use), alpha = 0.8) +
  annotate("text", x = -0.15, y = 320, label = r2_text, parse = TRUE, hjust = 0) +
  annotate("text", x = -0.15, y = 275, label = p_text, parse = TRUE, hjust = 0) + 
  labs(color = "Land use",
       x=expression("Growing season water table level (m)"), 
       y=expression("C accumulation rate " ~(g ~C ~m^-2 ~yr^-1))) +
  scale_color_manual(values = c("Reference" = "#D55E00", "Restored" = "#619CFF", "Disturbed" = "#009E73")) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 10),    
        axis.title = element_text(size = 10),   
        legend.text = element_text(size = 10),  
        legend.title = element_text(size = 10),
        legend.position = c(0.2, 0.8)) 
p6

```

### Lat vs CAR

```{r}
fit <- lm(car_all ~ lat, data = data)
r2 <- summary(fit)$r.squared
p_val <- summary(fit)$coefficients[2, 4]

r2_text <- paste0("italic(R)^2 == ", format(round(r2, 2), nsmall = 2))
p_text <- paste0("italic(p) == ", format(round(p_val, 2), nsmall = 3))

p7 <- ggplot(data, aes(x=lat, y=car_all)) +
        geom_smooth(method = "lm", se=TRUE, alpha = 0.4) +
  geom_point(aes(color = land_use), alpha = 0.8) +
  annotate("text", x = 47, y = 320, label = r2_text, parse = TRUE, hjust = 0) +
  annotate("text", x = 47, y = 275, label = p_text, parse = TRUE, hjust = 0) + 
  labs(color = "Land use",
       x=expression("Latitude"), 
       y=expression("C accumulation rate " ~(g ~C ~m^-2 ~yr^-1))) +
  scale_color_manual(values = c("Reference" = "#D55E00", "Restored" = "#619CFF", "Disturbed" = "#009E73")) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 10),  
        axis.title = element_text(size = 10),   
        legend.text = element_text(size = 10),  
        legend.title = element_text(size = 10),
        legend.position = c(0.2, 0.8)) 
p7

```

### Salinity vs CAR

```{r}
fit <- lm(car_all ~ sal_annual, data = data)
r2 <- summary(fit)$r.squared
p_val <- summary(fit)$coefficients[2, 4]

r2_text <- paste0("italic(R)^2 == ", format(round(r2, 2), nsmall = 2))
p_text <- paste0("italic(p) == ", format(round(p_val, 3), nsmall = 3))

p8 <- ggplot(data, aes(x=sal_annual, y=car_all)) +
        geom_smooth(method = "lm", se=TRUE, alpha = 0.4) +
  geom_point(aes(color = land_use), alpha = 0.8) +
  annotate("text", x = 20, y = 320, label = r2_text, parse = TRUE, hjust = 0) +
  annotate("text", x = 20, y = 275, label = p_text, parse = TRUE, hjust = 0) + 
  labs(color = "Land use",
       x=expression("Annual salinity (ppt)"), 
       y=expression("C accumulation rate " ~(g ~C ~m^-2 ~yr^-1))) +
  scale_color_manual(values = c("Reference" = "#D55E00", "Restored" = "#619CFF", "Disturbed" = "#009E73")) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 10),    # Adjust axis text size
        axis.title = element_text(size = 10),   # Adjust axis title size
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 10),
        legend.position = c(0.2, 0.8)) 
p8

```

### Dist chan vs CAR

```{r}
fit <- lm(car_all ~ dist_chan, data = data_tidal)
r2 <- summary(fit)$r.squared
p_val <- summary(fit)$coefficients[2, 4]

r2_text <- paste0("italic(R)^2 == ", format(round(r2, 2), nsmall = 2))
p_text <- paste0("italic(p) == ", format(round(p_val, 2), nsmall = 3))

p9 <- ggplot(data_tidal, aes(x=dist_chan, y=car_all)) +
        geom_smooth(method = "lm", se=TRUE, alpha = 0.4) +
  geom_point(aes(color = land_use), alpha = 0.8) +
  annotate("text", x = 52, y = 325, label = r2_text, parse = TRUE, hjust = 0) +
  annotate("text", x = 52, y = 284, label = p_text, parse = TRUE, hjust = 0) +
  labs(color = "Land use",
       x=expression("Distance to channel (m)"), 
       y=expression("C accumulation rate " ~(g ~C ~m^-2 ~yr^-1))) +
  scale_color_manual(values = c("Reference" = "#D55E00", "Restored" = "#619CFF", "Disturbed" = "#009E73")) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 10),    # Adjust axis text size
        axis.title = element_text(size = 10),   # Adjust axis title size
        legend.text = element_text(size = 10),  # Adjust legend text size
        legend.title = element_text(size = 10),
        legend.position = c(0.2, 0.8)) 
p9

```

### combine 4 plots

```{r}
p6 <- p6 + theme(legend.position = c(0.2, 0.65),
                 legend.key.height = unit(0.5, "cm")) +
  annotate("text", x = -0.98, y = Inf, label = "A", size = 6, fontface = "plain", vjust = 1.5, hjust = -0.2) 
p7 <- p7 + theme(legend.position = "none") +
  annotate("text", x = 43.3, y = Inf, label = "B", size = 6, fontface = "plain", vjust = 1.5, hjust = -0.2) 
p8 <- p8 + theme(legend.position = "none") +
  annotate("text", x = 0.6, y = Inf, label = "C", size = 6, fontface = "plain", vjust = 1.5, hjust = -0.2) 
p9 <- p9 + theme(legend.position = "none") +
  annotate("text", x = 1, y = Inf, label = "D", size = 6, fontface = "plain", vjust = 1.5, hjust = -0.2) 

CAR_scatter_combined <- ggarrange(p6, p7, p8, p9, ncol = 2, nrow = 2)

print(CAR_scatter_combined)

#ggsave("CAR_scatter_combined.png", CAR_scatter_combined, width = 8, height = 5.5, units = "in", dpi = 300)
```

## Fig S1 - SET vs Pb210 accretion

```{r}

p10 <- ggplot(data=data_S1, aes(x=set, y=pb210)) +
  geom_smooth(method="lm", alpha = 0.4) +
  geom_point(size=2, aes(color=land_use), show.legend=TRUE) +
  stat_regline_equation(label.y = 1.4, aes(label=paste(..eq.label..))) +
  stat_cor(label.y = 1.26, aes(label= ..rr.label..)) +
  xlim(c(-1,5)) +
  labs(
    color = "Land use",
    y=expression("Pb-210 accretion rate" ~ (cm ~ yr^-1)), 
    x=expression("SET elevation change rate" ~(cm ~yr^-1))) +
  scale_color_manual(values = c("Reference" = "#D55E00", "Restored" = "#619CFF", "Disturbed" = "#009E73")) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 10),    # Adjust axis text size
        axis.title = element_text(size = 10),   # Adjust axis title size
        legend.text = element_text(size = 9),  # Adjust legend text size
        legend.title = element_text(size = 10),
        legend.position = c(0.85, 0.22),
        legend.key.height = unit(0.5, "cm")) 

pb210_model <- lm(pb210 ~ set, data_S1)
summary(pb210_model)
p10

#ggsave("Pb210_vs_SET.png", plot = p10, height=3, width=5, units="in", dpi = 300)

```

## Fig S2 - Accretion vs CD

```{r}
p11 <- ggplot(data, aes(x=cd, y=pb210_all)) +
        geom_smooth(method = "lm", se=TRUE, alpha = 0.4) +
  geom_point(aes(color = land_use), alpha = 0.8) +
  stat_cor(method = "pearson", label.x = 11, label.y = 1.0, r.accuracy = 0.01, 
           aes(label = ..r.label..)) +
  labs(color = "Land use",
       x=expression("C density " ~(mg ~C ~cm^-3)), 
       y=expression("Accretion rate " ~(cm ~yr^-1))) +
  scale_color_manual(values = c("Reference" = "#D55E00", "Restored" = "#619CFF", "Disturbed" = "#009E73")) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 10),    # Adjust axis text size
        axis.title = element_text(size = 10),   # Adjust axis title size
        legend.text = element_text(size = 9),  # Adjust legend text size
        legend.title = element_text(size = 10),
        legend.position = c(0.85, 0.75),
        legend.key.height = unit(0.5, "cm")) 
p11

#ggsave("pb210_vs_CD.png", p11, width = 5, height = 3, units = "in", dpi = 300)

model <- lm(pb210_all ~ cd, data = data)
summary(model)

```



